Hydrodynamics of a Semipermeable Inextensible Membrane Under Flow and Confinement

### 2021-05-20 vesicle_code/docs/permeableVes/semipermeable_vesicle.tex
yny

The cellular membrane is a lipid bilayer membrane that is slightly permeable to water and lipid-soluble
molecules~\cite{Dick1964_JTB, FettiplaceHaydon1980_PhysRev,
DeamerBramhall1986_ChemPhysLipids, Grafmueller2019_ABLS}. 
%
%In the presence of an osmotic gradient across the membrane (such as a contrast in the concentration of an impermeable solute), water flows against the osmotic gradient  to neutralize the osmolarity. 
%
In the red blood cell membrane, aquaporin-1 channels give rise to a much
higher water permeability coefficient of approximately $1.8\times 10^{-2}$ cm/s in
erythrocytes~\cite{YangMaVerkman2001_JBC} while the intrinsic water
permeability of a lipid bilayer membrane is in the range of
$10^{-4}$--$10^{-3}$ cm/s~\cite{ThompsonHuang1966_ANYAS,
FettiplaceHaydon1980_PhysRev, Grafmueller2019_ABLS, Dimova2020_GVB,
BhatiaRobinsonDimova2020_SoftMatt}. 
Without aquaporins or other membrane channels to promote water
permeability, the intrinsic membrane permeability is known to depend on the
temperature, lipid composition~\cite{OlbrichRawiczNeedhamEtAl2000_BJ},
and the hydration of the membrane~\cite{MarrinkBerendsen1994_JPhysChem}.
%
%The swelling/shrinking of a cell is inevitably coupled to changes in
%membrane tension and deformation. Under a strong confinement (such as
%going through a microcapillary or submicron slit),  the transmembrane
%water flow may have comparable contributions from both osmotic and
%mechanical stresses
%\cite{StrokaJiangChenEtAl2014_Cell,LiMoriSun2015_PRL,yao-mor2017}.  
Molecular dynamic (MD) simulations show that a stretched red blood cell
membrane porates when the membrane tension reaches the order of $2$--$4$
mN/m (see~\citet{RazizadehNikfarPaulLiu2020_BJ} and references therein),
giving rise to enhanced membrane permeability to both water and
macromolecules such as ATP. Consequently, as long as the lipid bilayer membrane
remains intact (no poration or rupture), the total amount of water inside the membrane 
is assumed constant throughout the experiment of duration
no longer than tens of minutes.
%
%membrane poration is inevitable when membrane tension is of the order
%of $2-4$ mN/m (\cite{RazizadehNikfarPaulLiu2020_BJ} and references
%therein). 
%%The CGMD results further show that a stretched RBC membrane may have pores that might contribute to transmembrane transport of both water and ATP molecules \cite{RazizadehNikfarPaulLiu2020_BJ}. 
%Thus an important outstanding question in stress-induced ATP release of
%highly deformed RBCs is if there is any transmembrane flow when RBCs
%release ATP as they experience large mechanical load due to confinement
%or large shear stress?  To address such question, the first step we
%take is to focus on the permeable water flux induced by mechanical
%stress.
%%
%%RBCs are known to release ATP under large deformation and microfluidic experiments have been used to study the correlation between RBC deformation and ATP release \cite{Wan2008_PNAS,ForsythWan2011_PNAS,ZhangShenHoganBarakatMisbah2018_BJ}.
%%
%%Why is it interesting to study the dynamics of a semipermeable membrane induced by mechanical stress/confinement?
%%
%%The osmolarity-induced water flux has been well studied, while mechanical pressure may also contribute to the permeable water flow \cite{yao-mor2017}.
%%The response of a cell to an osmotic gradient is inevitably both mechanical (tension redistribution and shape change) and chemical (water flux across the membrane).  
%
%The cellular nuclear envelop (NE) consists of two lipid bilayer membranes connected via nuclear pore complexes (NPCs). As a result, NE is has a higher hydraulic conductivity
%than a lipid bilayer membrane. Even though the NPCs are sufficiently large for transport of both macromolecules and water molecules, the colloid-osmotic pressure inside
%the nucleus is highly effective in regulating the water content (see \citet{Shahin2019_AdvSci} and references therein).

@@

We non-dimensionalize the governing equations (Equations~(\ref{eqn:water_flux}-\ref{eqn:jump})) by a
characteristic length scale $R_0 = 10^{-6}$ m, a bending stiffness $k_b
= 10^{-19}$ J, and fluid viscosity $5 \times 10^{-2}$ kg/ms. We scale
time by $\tau = \mu R_0^3/k_b = 0.5\;s$, the velocity by $R_0/\tau =
2\;\mu$m/s, the pressure by $k_b/R_0^3 = 10^{-1}$ Pa, the tension by
$k_b/R_0^2 = 10^{-7}$ N/m, and the permeability by $R_0/\mu = 10^{-3}$
m$^2$s/kg. Under this nondimensionalization, a two-dimensional
semipermeable vesicle in a viscous fluid is characterized by (a) its
reduced area $\alpha = 4\pi A/L^2 \in (0,1]$, where $L$ is the vesicle
length and $A$ is its area, and (b) its permeability $\beta = k_w \mu
/R_0$. 
In two dimensions, the conservation of total lipids in the membrane corresponds to the inextensibility (fixed length $L$) of $\gamma$.
Due to semipermeability to water ($\beta>0$),  the enclosed area $A$ is not conserved and the reduced area $\alpha$ is dynamic.

>>>>

The cellular membrane is a lipid bilayer membrane that is slightly permeable to water and lipid-soluble
molecules~\cite{Dick1964_JTB, FettiplaceHaydon1980_PhysRev,
DeamerBramhall1986_ChemPhysLipids, Grafmueller2019_ABLS}. 
%
%In the presence of an osmotic gradient across the membrane (such as a contrast in the concentration of an impermeable solute), water flows against the osmotic gradient  to neutralize the osmolarity. 
%
In the red blood cell membrane, aquaporin-1 channels give rise to a much
higher water permeability coefficient of approximately $1.8\times 10^{-2}$ cm/s in
erythrocytes~\cite{YangMaVerkman2001_JBC} while the intrinsic water
permeability of a lipid bilayer membrane is in the range of
$10^{-4}$--$10^{-3}$ cm/s~\cite{ThompsonHuang1966_ANYAS,
FettiplaceHaydon1980_PhysRev, Grafmueller2019_ABLS, Dimova2020_GVB,
BhatiaRobinsonDimova2020_SoftMatt}. 
Without aquaporins or other membrane channels to promote water
permeability, the intrinsic membrane permeability is known to depend on the
temperature, lipid composition~\cite{OlbrichRawiczNeedhamEtAl2000_BJ},
and the hydration of the membrane~\cite{MarrinkBerendsen1994_JPhysChem}.
%
%The swelling/shrinking of a cell is inevitably coupled to changes in
%membrane tension and deformation. Under a strong confinement (such as
%going through a microcapillary or submicron slit),  the transmembrane
%water flow may have comparable contributions from both osmotic and
%mechanical stresses
%\cite{StrokaJiangChenEtAl2014_Cell,LiMoriSun2015_PRL,yao-mor2017}.  
Molecular dynamic (MD) simulations show that a stretched red blood cell
membrane porates when the membrane tension reaches the order of $2$--$4$
mN/m,
giving rise to \textcolor{magenta}{membrane rupture and leakage of macromolecules (see~\citet{RazizadehNikfarPaulLiu2020_BJ} and references therein).}
%
%enhanced membrane permeability to both water and
%macromolecules such as ATP. 
%
Consequently, as long as the lipid bilayer membrane
remains intact (no poration or rupture), the total amount of water inside the membrane 
is assumed constant throughout the experiment of duration
no longer than tens of minutes.
%
%membrane poration is inevitable when membrane tension is of the order
%of $2-4$ mN/m (\cite{RazizadehNikfarPaulLiu2020_BJ} and references
%therein). 
%%The CGMD results further show that a stretched RBC membrane may have pores that might contribute to transmembrane transport of both water and ATP molecules \cite{RazizadehNikfarPaulLiu2020_BJ}. 
%Thus an important outstanding question in stress-induced ATP release of
%highly deformed RBCs is if there is any transmembrane flow when RBCs
%release ATP as they experience large mechanical load due to confinement
%or large shear stress?  To address such question, the first step we
%take is to focus on the permeable water flux induced by mechanical
%stress.
%%
%%RBCs are known to release ATP under large deformation and microfluidic experiments have been used to study the correlation between RBC deformation and ATP release \cite{Wan2008_PNAS,ForsythWan2011_PNAS,ZhangShenHoganBarakatMisbah2018_BJ}.
%%
%%Why is it interesting to study the dynamics of a semipermeable membrane induced by mechanical stress/confinement?
%%
%%The osmolarity-induced water flux has been well studied, while mechanical pressure may also contribute to the permeable water flow \cite{yao-mor2017}.
%%The response of a cell to an osmotic gradient is inevitably both mechanical (tension redistribution and shape change) and chemical (water flux across the membrane).  
%
%The cellular nuclear envelop (NE) consists of two lipid bilayer membranes connected via nuclear pore complexes (NPCs). As a result, NE is has a higher hydraulic conductivity
%than a lipid bilayer membrane. Even though the NPCs are sufficiently large for transport of both macromolecules and water molecules, the colloid-osmotic pressure inside
%the nucleus is highly effective in regulating the water content (see \citet{Shahin2019_AdvSci} and references therein).

@@

We non-dimensionalize the governing equations (Equations~(\ref{eqn:water_flux}-\ref{eqn:jump})) by a
characteristic length scale $R_0 = 10^{-6}$ m, a bending stiffness $k_b
= 10^{-19}$ J, and fluid viscosity $5 \times 10^{-2}$ kg/ms. We scale
time by $\tau = \mu R_0^3/k_b = 0.5\;s$, the velocity by $R_0/\tau =
2\;\mu$m/s, the pressure by $k_b/R_0^3 = 10^{-1}$ Pa, the tension by
$k_b/R_0^2 = 10^{-7}$ N/m, and the permeability by $R_0/\mu = 10^{-3}$
m$^2$s/kg. Under this nondimensionalization \textcolor{magenta}{suitable for a vesicle in a quiescent flow}, a two-dimensional
semipermeable vesicle in a viscous fluid is characterized by (a) its
reduced area $\alpha = 4\pi A/L^2 \in (0,1]$, where $L$ is the vesicle
length and $A$ is its area, and (b) its permeability $\beta = k_w \mu
/R_0$. 
In two dimensions, the conservation of total lipids in the membrane corresponds to the inextensibility (fixed length $L$) of $\gamma$.
Due to semipermeability to water ($\beta>0$),  the enclosed area $A$ is not conserved and the reduced area $\alpha$ is dynamic.
